---
title: "Exploratory Data Analysis"
subtitle: "Looking before we leap — the patterns that motivate our features"
---

> **Where we are:** This is step one of the
> [Titanic — Tidymodels Tour](../index.qmd). Before fitting any model we look at
> the data, find the signal, and let what we see *motivate* the shared feature
> recipe in `R/02_features.R` (`make_recipe()`). Every model document downstream
> trains on those features, so the decisions we justify here ripple through the
> whole project.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

# Source the project setup: loads tidyverse + tidymodels, sets seed = 42,
# defines path_* helpers, and registers the parallel backend.
source(here::here("R", "00_setup.R"))

# Attach the libraries whose functions are used bare throughout this doc.
library(tidyverse)   # dplyr, ggplot2, readr, stringr, tidyr, purrr, forcats, etc.
library(tidymodels)  # recipes and friends for the shared feature recipe
library(here)        # here() for project-relative paths

# A couple of EDA-only helpers.
library(skimr)   # skim() — richer than summary()
library(naniar)  # missingness visualisation (optional but lovely)
library(scales)  # percent()/dollar() axis + label formatting (not attached by tidyverse)
```

```{r}
#| label: load-data
#| message: false

# Read the training data. Only the training set has the Survived label, so all
# exploratory analysis happens here.
train <- read_csv(here("data", "raw", "train.csv"))

# Survived is the outcome. Treat it as a factor with explicit levels.
# Convention for the whole project: levels c("0", "1"), positive class = "1".
train <- train |>
  mutate(Survived = factor(Survived, levels = c("0", "1")))

glimpse(train)
```

## First look: structure and missingness

```{r}
#| label: skim

# skim() gives per-type summaries: completeness, mean/sd for numerics,
# top counts for characters. The `complete_rate` column is the one to watch.
skim(train)
```

The headline structural facts:

- **891 passengers**, the outcome `Survived` is binary.
- **`Age` is missing for ~20%** of passengers — too much to drop, so we will
  *impute* it in the recipe.
- **`Cabin` is missing for ~77%** — mostly unusable as-is, though its
  *presence/absence* may carry signal (a recorded cabin ≈ higher class).
- **`Embarked` has 2 missing values** — trivially imputable with the mode.

```{r}
#| label: missingness
#| fig-cap: "Proportion of missing values per column. Age and Cabin dominate."
#| fig-width: 7
#| fig-height: 4

# A quick ggplot of missingness so we can *see* what skim told us.
train |>
  summarise(across(everything(), ~ mean(is.na(.x)))) |>
  pivot_longer(everything(),
                      names_to = "variable", values_to = "pct_missing") |>
  filter(pct_missing > 0) |>
  ggplot(aes(x = reorder(variable, pct_missing),
                               y = pct_missing)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = percent(pct_missing, accuracy = 1)),
                     hjust = -0.1, size = 3.5) +
  scale_y_continuous(labels = percent,
                              expand = expansion(mult = c(0, 0.15))) +
  coord_flip() +
  labs(title = "Missingness by column",
                x = NULL, y = "Missing") +
  theme_minimal()
```

The overall survival rate is our **base rate** — the number to beat with any
model:

```{r}
#| label: base-rate
train |>
  count(Survived) |>
  mutate(prop = n / sum(n))
```

About **38%** survived. A "predict everyone died" baseline would therefore score
~62% accuracy — keep that in mind when judging model scores.

## Who survived? The classic patterns

### Survival by sex

```{r}
#| label: survival-sex
#| fig-cap: "Survival rate by sex. The single strongest predictor in the data."
#| fig-width: 6
#| fig-height: 4

train |>
  group_by(Sex) |>
  summarise(survival_rate = mean(Survived == "1"), n = n()) |>
  ggplot(aes(x = Sex, y = survival_rate, fill = Sex)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = percent(survival_rate, accuracy = 1)),
                     vjust = -0.4) +
  scale_y_continuous(labels = percent,
                              limits = c(0, 1)) +
  labs(title = "\"Women and children first\"",
                x = NULL, y = "Survival rate") +
  theme_minimal() +
  theme(legend.position = "none")
```

**~74% of women survived vs ~19% of men.** `Sex` is the dominant predictor —
any model that ignores it is leaving the table.

### Survival by passenger class

```{r}
#| label: survival-pclass
#| fig-cap: "Survival rate by passenger class (1 = first class)."
#| fig-width: 6
#| fig-height: 4

train |>
  mutate(Pclass = factor(Pclass)) |>
  group_by(Pclass) |>
  summarise(survival_rate = mean(Survived == "1")) |>
  ggplot(aes(x = Pclass, y = survival_rate, fill = Pclass)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = percent(survival_rate, accuracy = 1)),
                     vjust = -0.4) +
  scale_y_continuous(labels = percent, limits = c(0, 1)) +
  labs(title = "Class mattered",
                x = "Passenger class", y = "Survival rate") +
  theme_minimal() +
  theme(legend.position = "none")
```

A clear gradient: **1st class ~63%, 2nd ~47%, 3rd ~24%.** Class is a proxy for
deck location, wealth, and access to lifeboats.

### Sex and class together

```{r}
#| label: survival-sex-pclass
#| fig-cap: "Survival by sex within each class — the interaction is striking."
#| fig-width: 7
#| fig-height: 4

train |>
  mutate(Pclass = factor(Pclass)) |>
  group_by(Sex, Pclass) |>
  summarise(survival_rate = mean(Survived == "1"), .groups = "drop") |>
  ggplot(aes(x = Pclass, y = survival_rate, fill = Sex)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_y_continuous(labels = percent, limits = c(0, 1)) +
  labs(title = "Sex × class interaction",
                x = "Passenger class", y = "Survival rate") +
  theme_minimal()
```

First-class women survived almost universally; third-class men almost never.
Tree-based models capture this interaction automatically; for the neural net it
helps to have rich, normalized features.

## Age, fare, and continuous structure

### Age distribution by survival

```{r}
#| label: age-density
#| fig-cap: "Age distribution split by survival. Note the spike of saved children."
#| fig-width: 7
#| fig-height: 4
#| warning: false

train |>
  ggplot(aes(x = Age, fill = Survived)) +
  geom_histogram(binwidth = 5, position = "identity", alpha = 0.6) +
  scale_fill_manual(values = c("0" = "grey50", "1" = "seagreen"),
                             labels = c("Died", "Survived")) +
  labs(title = "Age distribution by outcome",
                x = "Age (years)", y = "Count", fill = NULL) +
  theme_minimal()
```

```{r}
#| label: age-density-2
#| fig-cap: "Same story as a density — young children survived disproportionately."
#| fig-width: 7
#| fig-height: 4
#| warning: false

train |>
  ggplot(aes(x = Age, fill = Survived)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("0" = "grey50", "1" = "seagreen"),
                             labels = c("Died", "Survived")) +
  labs(title = "Age density by outcome",
                x = "Age (years)", y = "Density", fill = NULL) +
  theme_minimal()
```

Children (roughly under 10) survived at a higher rate — the "and children"
half of the rule. Because `Age` is ~20% missing, **imputation** in the recipe
is essential; a median or bagged-tree imputation keeps these rows in play.

### Fare by class

```{r}
#| label: fare-class
#| fig-cap: "Fare by class (log scale). Fare is largely a continuous echo of class."
#| fig-width: 7
#| fig-height: 4
#| warning: false

train |>
  mutate(Pclass = factor(Pclass)) |>
  ggplot(aes(x = Pclass, y = Fare, fill = Pclass)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  scale_y_log10(labels = dollar) +
  labs(title = "Fare by passenger class (log scale)",
                x = "Passenger class", y = "Fare (log10)") +
  theme_minimal() +
  theme(legend.position = "none")
```

`Fare` is right-skewed and strongly tied to class — a candidate for a **log
transform** so models (especially the neural net) see a better-behaved
distribution.

## Engineered signal: family and title

These two features are not columns in the raw data — we *derive* them. This is
the heart of feature engineering and exactly what `make_recipe()` will encode.

### Family size

`SibSp` (siblings/spouses aboard) + `Parch` (parents/children aboard) + 1 (self)
gives a **family size**. Travelling alone vs. in a small vs. large group changes
survival odds.

```{r}
#| label: family-size
#| fig-cap: "Survival rate by family size. Small families fared best."
#| fig-width: 7
#| fig-height: 4

train |>
  mutate(FamilySize = SibSp + Parch + 1) |>
  group_by(FamilySize) |>
  summarise(survival_rate = mean(Survived == "1"), n = n(),
                   .groups = "drop") |>
  ggplot(aes(x = FamilySize, y = survival_rate)) +
  geom_col(aes(fill = n), width = 0.7) +
  geom_text(aes(label = n), vjust = -0.4, size = 3) +
  scale_y_continuous(labels = percent, limits = c(0, 1)) +
  scale_fill_viridis_c(name = "n passengers") +
  scale_x_continuous(breaks = 1:11) +
  labs(title = "Survival by family size",
                subtitle = "Solo travellers and large families did worst",
                x = "Family size (SibSp + Parch + 1)", y = "Survival rate") +
  theme_minimal()
```

A clear "Goldilocks" pattern: **family size 2–4 survived best**, while solo
travellers and very large families fared worst. This motivates both a continuous
`FamilySize` feature and possibly a binned `IsAlone` flag.

### Title extracted from name

The `Name` column embeds a **title** (Mr, Mrs, Miss, Master, and rarer ones like
Dr, Rev, Col). Title bundles sex, age band, and social status into one categorical
feature — and `Master` in particular flags young boys, who survived more often
than adult men.

```{r}
#| label: title
#| fig-cap: "Survival rate by extracted title (titles with < 10 passengers grouped as 'Rare')."
#| fig-width: 8
#| fig-height: 4

train |>
  # Pull the title: the word ending in "." after the comma in the Name field.
  mutate(
    Title = str_extract(Name, "(?<=, )[A-Za-z]+(?=\\.)")
  ) |>
  # Collapse the long tail of rare titles so the plot (and later the model)
  # is not dominated by singletons.
  add_count(Title) |>
  mutate(Title = if_else(n < 10, "Rare", Title)) |>
  group_by(Title) |>
  summarise(survival_rate = mean(Survived == "1"), n = n(),
                   .groups = "drop") |>
  ggplot(aes(x = reorder(Title, survival_rate),
                               y = survival_rate)) +
  geom_col(aes(fill = n), width = 0.7) +
  geom_text(aes(label = n), hjust = -0.1, size = 3) +
  scale_y_continuous(labels = percent,
                              limits = c(0, 1.05),
                              expand = expansion(mult = c(0, 0.05))) +
  scale_fill_viridis_c(name = "n passengers") +
  coord_flip() +
  labs(title = "Survival by title",
                x = NULL, y = "Survival rate") +
  theme_minimal()
```

`Mrs`/`Miss` survived at high rates, `Master` (boys) noticeably better than
`Mr`. Title is one of the most valuable engineered features in this dataset.

## What this means for `make_recipe()`

Everything above is the *justification* for the shared recipe in
`R/02_features.R`. In words, `make_recipe(data)` should:

1. **Impute `Age`** (missing ~20%) — median or bagged imputation.
2. **Impute `Embarked`** with the mode (only 2 missing).
3. **Derive `FamilySize`** = `SibSp + Parch + 1`, and optionally an `IsAlone`
   flag.
4. **Extract `Title`** from `Name` and lump rare titles together.
5. **Log-transform `Fare`** (right-skewed) and consider a `HasCabin` flag from
   `Cabin` presence.
6. **Encode factors** (`Sex`, `Embarked`, `Pclass`, `Title`) as dummies for
   models that need numeric input.
7. **Drop identifiers** (`PassengerId`, `Ticket`, raw `Name`, raw `Cabin`).

::: {.callout-note}
The neural-net document adds one more step on top of the shared recipe —
`step_normalize(all_numeric_predictors())` — because MLPs are sensitive to input
scale in a way that tree ensembles are not. See
[the Neural Net doc](04-neural-net.qmd) for why.
:::

## Next steps

You now know *why* the features look the way they do. Time to model:

- **[XGBoost](02-xgboost.qmd)** — start here; it's the Kaggle workhorse and
  establishes the modelling template the other docs reuse.
- **[Random Forest](03-random-forest.qmd)** — a robust, low-tuning comparison.
- **[Neural Net](04-neural-net.qmd)** — a small MLP that leans on the normalized
  recipe.
- Back to the **[master document](../index.qmd)** for the big picture and the
  results table.
