---
title: "Model 1 — XGBoost"
subtitle: "Gradient-boosted trees, the Kaggle workhorse"
---

> **Where we are:** First of three model documents in the
> [Titanic — Tidymodels Tour](../index.qmd). All three follow the *same*
> structure — recipe → spec → workflow → tune → finalize → evaluate → save — so
> once you've read this one, [Random Forest](03-random-forest.qmd) and
> [Neural Net](04-neural-net.qmd) are quick to skim for the differences. The
> features come from the [EDA](01-eda.qmd) and the shared `make_recipe()`.

## Why XGBoost first?

Gradient-boosted decision trees are the default first reach on tabular Kaggle
problems. They handle mixed numeric/categorical features, capture interactions
(like sex × class) automatically, are robust to monotone transforms, and tune to
strong accuracy with modest effort. We tune four of the most impactful knobs.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

# Project setup: tidyverse + tidymodels, seed = 42, paths, parallel backend.
source(here::here("R", "00_setup.R"))
# Shared feature engineering: defines make_recipe(data).
source(here("R", "02_features.R"))

# Attach the libraries used throughout this document.
library(tidyverse)   # dplyr, ggplot2, readr, stringr, tibble, tidyr, purrr
library(tidymodels)  # recipes, parsnip, workflows, tune, yardstick, rsample, dials, broom
library(here)        # path construction
library(vip)         # variable importance plots
library(xgboost)     # the engine
```

```{r}
#| label: load-data
#| message: false

train <- readr::read_csv(here("data", "processed", "train.csv")) |>
  # Project-wide convention: Survived is a factor, levels c("0","1"),
  # positive class = "1" (so metrics use event_level = "second").
  mutate(Survived = factor(Survived, levels = c("0", "1")))
```

## The recipe

The shared recipe encodes everything we justified in the
[EDA](01-eda.qmd): impute `Age`/`Embarked`, derive `FamilySize`, extract
`Title`, log-transform `Fare`, dummy-encode factors, drop identifiers.

```{r}
#| label: recipe
xgb_recipe <- make_recipe(train)
xgb_recipe
```

## The model specification

We mark the four tuning parameters with `tune()`. tidymodels will fill them in
during the grid search.

```{r}
#| label: spec
xgb_spec <- boost_tree(
  trees      = tune(),
  tree_depth = tune(),
  learn_rate = tune(),
  min_n      = tune()
) |>
  set_engine("xgboost") |>
  set_mode("classification")

xgb_spec
```

## The workflow

A `workflow` bundles the recipe and the model so resampling applies the *exact*
same preprocessing to every fold — no leakage.

```{r}
#| label: workflow
xgb_wf <- workflow() |>
  add_recipe(xgb_recipe) |>
  add_model(xgb_spec)
```

## Tuning with 5-fold cross-validation

```{r}
#| label: resamples
# Stratify on the outcome so each fold preserves the ~38% survival rate.
set.seed(42)
folds <- vfold_cv(train, v = 10, strata = Survived)
```

```{r}
#| label: grid
# Predictors used to finalize data-dependent params (mtry, num_comp upper bound).
preds <- prep(make_recipe(train, num_comp = 2)) |>
  bake(new_data = NULL) |>
  select(-Survived, -PassengerId)

xgb_params <- extract_parameter_set_dials(xgb_wf) |>
  finalize(preds) |>
  update(num_comp = num_comp(range = c(2L, 10L)))

set.seed(42)
xgb_grid <- grid_space_filling(xgb_params, size = 18)
```

```{r}
#| label: tune
#| message: false
#| warning: false

# Score every candidate on both metrics. event_level = "second" because the
# positive class ("survived" = "1") is the SECOND factor level.
xgb_metrics <- metric_set(roc_auc, accuracy)

set.seed(42)
xgb_tuned <- tune_grid(
  xgb_wf,
  resamples = folds,
  grid      = xgb_grid,
  metrics   = xgb_metrics,
  control   = control_grid(save_pred = TRUE,
                           save_workflow = TRUE,
                           event_level = "second")
)
```

```{r}
#| label: save-res
saveRDS(xgb_tuned, here("models", "xgboost_res.rds"))
```

### Tuning results

```{r}
#| label: tune-plot
#| fig-cap: "ROC-AUC across the tuning grid for each hyperparameter (autoplot)."
#| fig-width: 9
#| fig-height: 6

autoplot(xgb_tuned, metric = "roc_auc")
```

```{r}
#| label: tune-table
# The top candidates by ROC-AUC.
show_best(xgb_tuned, metric = "roc_auc", n = 5)
```

## Finalize and fit on the full training set

```{r}
#| label: finalize
# Pick the best ROC-AUC candidate and lock those hyperparameters in.
best_xgb <- select_best(xgb_tuned, metric = "roc_auc")
best_xgb

final_xgb_wf <- finalize_workflow(xgb_wf, best_xgb)

# Fit the finalized workflow on ALL the training data — this is the model we
# will ship to Kaggle.
set.seed(42)
xgb_fit <- fit(final_xgb_wf, data = train)
```

## Cross-validated performance

The fairest estimate of out-of-sample skill is the CV score of the chosen
configuration — not the fit on the full training data (which overfits by
construction).

```{r}
#| label: cv-metrics
# Collect the CV metrics for the winning hyperparameters.
collect_metrics(xgb_tuned) |>
  semi_join(best_xgb, by = names(best_xgb)) |>
  select(.metric, mean, std_err, n)
```

::: {.callout-tip}
Copy the `accuracy` and `roc_auc` `mean` values into the results table on the
[master document](../index.qmd).
:::

### Confusion matrix

Built from the held-out CV predictions for the best configuration — so it
reflects generalization, not memorization.

```{r}
#| label: confusion
#| fig-cap: "Confusion matrix from out-of-fold CV predictions."
#| fig-width: 5
#| fig-height: 4

xgb_cv_preds <- collect_predictions(xgb_tuned, parameters = best_xgb)

xgb_cv_preds |>
  conf_mat(truth = Survived, estimate = .pred_class) |>
  autoplot(type = "heatmap") +
  labs(title = "XGBoost — CV confusion matrix")
```

### Variable importance

```{r}
#| label: vip
#| fig-cap: "Top features by XGBoost gain. Title, Sex, Fare and class dominate."
#| fig-width: 7
#| fig-height: 5

xgb_fit |>
  extract_fit_parsnip() |>
  vip(num_features = 15, geom = "col",
      aesthetics = list(fill = "steelblue")) +
  labs(title = "XGBoost variable importance") +
  theme_minimal()
```

The engineered `Title`, plus `Sex`, `Fare`, and class, sit at the top — exactly
the signal the [EDA](01-eda.qmd) surfaced. Feature engineering earns its keep.

## Save the fitted workflow

```{r}
#| label: save
# Persist the full fitted workflow (recipe + model) so the submission document
# can load it without refitting. Project convention: models/<model>_fit.rds.
saveRDS(xgb_fit, file = here("models", "xgboost_fit.rds"))
```

## Submission

Generate this model's leaderboard CSV with the **parameterized**
[submission document](submission.qmd):

```bash
quarto render docs/submission.qmd -P model:xgboost
```

That loads `models/xgboost_fit.rds`, predicts on `data/raw/test.csv`, writes
`models/submission_xgboost.csv` (columns `PassengerId`, `Survived` as 0/1
integers), and prints the exact `kaggle competitions submit` command. See the
[Kaggle setup guide](setup-kaggle.md) for authentication.

## Next steps

- Compare against the **[Random Forest](03-random-forest.qmd)** — a robust,
  almost-no-tuning baseline.
- Then the **[Neural Net](04-neural-net.qmd)**, which needs normalized inputs.
- Build the CSV with the **[Submission doc](submission.qmd)**.
- Record this model's CV scores in the **[master results table](../index.qmd)**.
