---
title: "Model 2 — Random Forest"
subtitle: "Bagged trees with ranger — a robust, low-tuning baseline"
---

> **Where we are:** Second of three model documents in the
> [Titanic — Tidymodels Tour](../index.qmd). It follows the *same* structure as
> the [XGBoost doc](02-xgboost.qmd) — recipe → spec → workflow → tune →
> finalize → evaluate → save — so skim for the differences. Features come from
> the [EDA](01-eda.qmd) via the shared `make_recipe()`.

## Why a random forest?

Where XGBoost grows trees *sequentially* to correct residuals, a random forest
grows many *independent* deep trees on bootstrap samples and averages them. It
is famously forgiving: with `trees = 1000` fixed, there are really only two
knobs worth tuning (`mtry`, `min_n`), it rarely overfits badly, and it gives a
strong baseline with almost no fuss. We use the fast `ranger` engine and ask it
for **permutation** importance.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

source(here::here("R", "00_setup.R"))      # tidyverse + tidymodels, seed = 42
source(here::here("R", "02_features.R"))   # make_recipe(data)

library(tidyverse)    # dplyr, ggplot2, readr, etc.
library(tidymodels)   # recipes, parsnip, workflows, tune, yardstick, rsample, dials
library(here)         # project-root-relative paths
library(vip)          # variable importance plots
library(ranger)       # the engine
```

```{r}
#| label: load-data
#| message: false

train <- read_csv(here("data", "processed", "train.csv")) |>
  mutate(Survived = factor(Survived, levels = c("0", "1")))
```

## The recipe

Identical shared recipe — that is the whole point: a fair comparison across
models on the same features.

```{r}
#| label: recipe
rf_recipe <- make_recipe(train)
rf_recipe
```

## The model specification

`trees` is fixed at 1000 (more is rarely worth it); we tune `mtry` and `min_n`.
We set `importance = "permutation"` so we can plot variable importance later.

```{r}
#| label: spec
  rf_spec <- rand_forest(mtry = tune(),
                         min_n = tune(),
                         trees = 1000) |>
    set_engine("ranger", importance = "permutation") |>
    set_mode("classification")


rf_spec
```

## The workflow

```{r}
#| label: workflow
rf_wf <- workflow() |>
  add_recipe(rf_recipe) |>
  add_model(rf_spec)
```

## Tuning with 5-fold cross-validation

```{r}
#| label: resamples
set.seed(42)
folds <- vfold_cv(train, v = 10, strata = Survived)
```

```{r}
#| label: grid
# mtry's upper bound depends on the number of predictors AFTER the recipe runs,
# so we finalize() its range against a prepped predictor set rather than
# guessing. num_comp (PCA) is tuned over a sensible range.
preds <- prep(make_recipe(train, num_comp = 2)) |>
  bake(new_data = NULL) |>
  select(-Survived, -PassengerId)

rf_params <- extract_parameter_set_dials(rf_wf) |>
  finalize(preds) |>
  update(num_comp = num_comp(range = c(2L, 10L)))

set.seed(42)
rf_grid <- grid_space_filling(rf_params, size = 18)
```

```{r}
#| label: tune
#| message: false
#| warning: false

rf_metrics <- metric_set(roc_auc, accuracy)

set.seed(42)
rf_tuned <- tune_grid(
  rf_wf,
  resamples = folds,
  grid      = rf_grid,
  metrics   = rf_metrics,
  # Positive class "1" is the second factor level.
  control   = control_grid(save_pred = TRUE, save_workflow = TRUE, event_level = "second")
)
```

```{r}
#| label: save-res
saveRDS(rf_tuned, here("models", "rf_res.rds"))
```

### Tuning results

```{r}
#| label: tune-plot
#| fig-cap: "ROC-AUC across the mtry / min_n grid (autoplot)."
#| fig-width: 8
#| fig-height: 5

autoplot(rf_tuned, metric = "roc_auc")
```

```{r}
#| label: tune-table
show_best(rf_tuned, metric = "roc_auc", n = 5)
```

## Finalize and fit on the full training set

```{r}
#| label: finalize
best_rf <- select_best(rf_tuned, metric = "roc_auc")
best_rf

final_rf_wf <- finalize_workflow(rf_wf, best_rf)

set.seed(42)
rf_fit <- fit(final_rf_wf, data = train)
```

## Cross-validated performance

```{r}
#| label: cv-metrics
collect_metrics(rf_tuned) |>
  semi_join(best_rf, by = names(best_rf)) |>
  select(.metric, mean, std_err, n)
```

::: {.callout-tip}
Copy the `accuracy` and `roc_auc` `mean` values into the results table on the
[master document](../index.qmd).
:::

### Confusion matrix

```{r}
#| label: confusion
#| fig-cap: "Confusion matrix from out-of-fold CV predictions."
#| fig-width: 5
#| fig-height: 4

rf_cv_preds <- collect_predictions(rf_tuned, parameters = best_rf)

rf_cv_preds |>
  conf_mat(truth = Survived, estimate = .pred_class) |>
  autoplot(type = "heatmap") +
  labs(title = "Random Forest — CV confusion matrix")
```

### Variable importance

```{r}
#| label: vip
#| fig-cap: "Permutation importance — drop in accuracy when each feature is shuffled."
#| fig-width: 7
#| fig-height: 5

rf_fit |>
  extract_fit_parsnip() |>
  vip(num_features = 15, geom = "col",
      aesthetics = list(fill = "forestgreen")) +
  labs(title = "Random Forest variable importance (permutation)") +
  theme_minimal()
```

Reassuringly, the forest leans on the same engineered features as XGBoost —
`Title`, `Sex`, `Fare`, class. Different algorithm, same story from the data.

## Save the fitted workflow

```{r}
#| label: save
# Project convention: models/<model>_fit.rds — note the file is rf_fit.rds.
saveRDS(rf_fit, file = here("models", "rf_fit.rds"))
```

## Submission

Generate this model's leaderboard CSV with the
[parameterized submission document](submission.qmd):

```bash
quarto render docs/submission.qmd -P model:rf
```

That loads `models/rf_fit.rds`, predicts on `data/raw/test.csv`, writes
`models/submission_rf.csv`, and prints the exact `kaggle competitions submit`
command. See the [Kaggle setup guide](setup-kaggle.md) for authentication.

## Next steps

- Finish the tour with the **[Neural Net](04-neural-net.qmd)** (needs normalized
  inputs).
- Revisit **[XGBoost](02-xgboost.qmd)** to compare importance and CV scores.
- Build the CSV with the **[Submission doc](submission.qmd)**.
- Record this model's CV scores in the **[master results table](../index.qmd)**.
