---
title: "Model 3 — Neural Net"
subtitle: "A small multilayer perceptron with brulee (torch) — and why scale matters"
---

> **Where we are:** Last of three model documents in the
> [Titanic — Tidymodels Tour](../index.qmd). Same structure as the
> [XGBoost](02-xgboost.qmd) and [Random Forest](03-random-forest.qmd) docs —
> recipe → spec → workflow → tune → finalize → evaluate → save. The one twist:
> a neural net needs **normalized** inputs, so this doc extends the shared
> `make_recipe()` with a normalization step.

## Why a neural net here?

On a dataset this small (891 rows) a multilayer perceptron will not beat the
tree ensembles — and that's fine. The point is to *explore* a different model
family and see how tidymodels makes swapping engines almost free. A small MLP
with a single hidden layer, weight decay (`penalty`), and a tuned number of
training `epochs` is a clean, instructive choice.

::: {.callout-note}
## brulee needs torch
The `brulee` engine fits neural nets via [torch](https://torch.mlverse.org/) for
R. The first time you use it, torch downloads its C++/LibTorch backend
(`torch::install_torch()` if it didn't auto-install). If torch is unavailable in
your environment (e.g. a locked-down CI runner), use the **nnet fallback** noted
below — the rest of the document is unchanged.
:::

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

source(here::here("R", "00_setup.R"))      # tidyverse + tidymodels, seed = 42
source(here::here("R", "02_features.R"))   # make_recipe(data)

library(brulee)   # the engine (torch-backed). See fallback note below.
library(vip)      # variable importance plots
```

```{r}
#| label: load-data
#| message: false

train <- readr::read_csv(here::here("data", "raw", "train.csv")) |>
  dplyr::mutate(Survived = factor(Survived, levels = c("0", "1")))
```

## The recipe — with normalization

Tree ensembles are invariant to monotone rescaling, so the shared recipe didn't
bother normalizing. A neural net is **not** invariant: gradient descent behaves
far better when every numeric predictor has mean 0 and unit variance, otherwise
large-scale features (like `Fare`) dominate the gradient. So we take the shared
recipe and append `step_normalize()`.

```{r}
#| label: recipe
# Extend the shared recipe with normalization of all numeric predictors. This
# step must come AFTER dummy encoding in make_recipe() so the 0/1 dummy columns
# are normalized too (standard practice for MLPs).
nnet_recipe <- make_recipe(train) |>
  step_normalize(all_numeric_predictors())

nnet_recipe
```

## The model specification

```{r}
#| label: spec
# brulee MLP: one hidden layer (hidden_units), L2 weight decay (penalty), and a
# tuned number of training epochs.
nnet_spec <- mlp(
  hidden_units = tune(),
  penalty      = tune(),
  epochs       = tune()
) |>
  set_engine("brulee") |>
  set_mode("classification")

nnet_spec
```

::: {.callout-important}
## Fallback: no torch? Use nnet
If `brulee`/`torch` cannot be installed, swap the engine to base R's `nnet`.
`nnet::nnet` does not expose an `epochs` argument the same way, so drop it and
tune just two parameters:

```r
nnet_spec <- mlp(
  hidden_units = tune(),
  penalty      = tune()
) |>
  set_engine("nnet") |>
  set_mode("classification")
```

Everything downstream (workflow, tuning, evaluation, save) works unchanged — just
remember to drop `epochs` from the tuning grid and from any `select_best()`
join keys below.
:::

## The workflow

```{r}
#| label: workflow
nnet_wf <- workflow() |>
  add_recipe(nnet_recipe) |>
  add_model(nnet_spec)
```

## Tuning with 5-fold cross-validation

```{r}
#| label: resamples
set.seed(42)
folds <- vfold_cv(train, v = 5, strata = Survived)
```

```{r}
#| label: grid
# Modest ranges suited to a tiny dataset: a handful of hidden units, light-to-
# moderate weight decay, and enough epochs to converge without overfitting.
set.seed(42)
nnet_grid <- grid_latin_hypercube(
  hidden_units(range = c(4L, 32L)),
  penalty(range = c(-4, -0.5)),     # log10 scale → 1e-4 to ~0.3
  epochs(range = c(50L, 300L)),
  size = 20
)
```

```{r}
#| label: tune
#| message: false
#| warning: false

nnet_metrics <- metric_set(roc_auc, accuracy)

set.seed(42)
nnet_tuned <- tune_grid(
  nnet_wf,
  resamples = folds,
  grid      = nnet_grid,
  metrics   = nnet_metrics,
  # Positive class "1" is the second factor level.
  control   = control_grid(save_pred = TRUE, event_level = "second")
)
```

### Tuning results

```{r}
#| label: tune-plot
#| fig-cap: "ROC-AUC across the MLP tuning grid (autoplot)."
#| fig-width: 9
#| fig-height: 5

autoplot(nnet_tuned, metric = "roc_auc")
```

```{r}
#| label: tune-table
show_best(nnet_tuned, metric = "roc_auc", n = 5)
```

## Finalize and fit on the full training set

```{r}
#| label: finalize
best_nnet <- select_best(nnet_tuned, metric = "roc_auc")
best_nnet

final_nnet_wf <- finalize_workflow(nnet_wf, best_nnet)

set.seed(42)
nnet_fit <- fit(final_nnet_wf, data = train)
```

## Cross-validated performance

```{r}
#| label: cv-metrics
# Note: if using the nnet fallback, drop "epochs" from the join keys.
collect_metrics(nnet_tuned) |>
  dplyr::semi_join(best_nnet,
                   by = c("hidden_units", "penalty", "epochs")) |>
  dplyr::select(.metric, mean, std_err, n)
```

::: {.callout-tip}
Copy the `accuracy` and `roc_auc` `mean` values into the results table on the
[master document](../index.qmd).
:::

### Confusion matrix

```{r}
#| label: confusion
#| fig-cap: "Confusion matrix from out-of-fold CV predictions."
#| fig-width: 5
#| fig-height: 4

nnet_cv_preds <- collect_predictions(nnet_tuned, parameters = best_nnet)

nnet_cv_preds |>
  conf_mat(truth = Survived, estimate = .pred_class) |>
  autoplot(type = "heatmap") +
  ggplot2::labs(title = "Neural Net — CV confusion matrix")
```

### Variable importance

Neural nets don't expose importance natively, so we use a **permutation-based**
importance via `vip` — shuffle each feature and measure the drop in ROC-AUC.

```{r}
#| label: vip
#| fig-cap: "Permutation importance for the MLP (drop in ROC-AUC when shuffled)."
#| fig-width: 7
#| fig-height: 5
#| message: false
#| warning: false

# A small prediction wrapper for vip's model-agnostic permutation method.
pred_wrapper <- function(object, newdata) {
  predict(object, new_data = newdata, type = "prob")$.pred_1
}

# Bake the (normalized) predictors for the permutation routine.
nnet_baked <- nnet_recipe |> prep() |> bake(new_data = train)

vip::vip(
  nnet_fit,
  method      = "permute",
  train       = nnet_baked,
  target      = "Survived",
  metric      = "roc_auc",
  event_level = "second",
  reference_class = "1",
  pred_wrapper = pred_wrapper,
  num_features = 15
) +
  ggplot2::labs(title = "Neural Net variable importance (permutation)") +
  ggplot2::theme_minimal()
```

Even through a different lens, the same features carry the weight — `Title`,
`Sex`, class, `Fare` — consistent with [XGBoost](02-xgboost.qmd) and the
[Random Forest](03-random-forest.qmd).

## Save the fitted workflow

```{r}
#| label: save
# Project convention: the neural-net model saves to nnet_fit.rds.
saveRDS(nnet_fit, file = here::here("models", "nnet_fit.rds"))
```

## Submission

Generate this model's leaderboard CSV with the
[parameterized submission document](submission.qmd):

```bash
quarto render docs/submission.qmd -P model:nnet
```

That loads `models/nnet_fit.rds`, predicts on `data/raw/test.csv`, writes
`models/submission_nnet.csv`, and prints the exact `kaggle competitions submit`
command. See the [Kaggle setup guide](setup-kaggle.md) for authentication.

## Next steps

- You've now fit all three families — compare their CV scores side by side in
  the **[master results table](../index.qmd)**.
- Revisit **[XGBoost](02-xgboost.qmd)** and the **[Random Forest](03-random-forest.qmd)**.
- Build the CSV with the **[Submission doc](submission.qmd)** and push it to the
  leaderboard.
