---
title: "Model 4 — Stacked Ensemble"
subtitle: "Blending XGBoost + Random Forest + Neural Net + Logistic Regression with stacks"
---

> **Where we are:** The capstone of the
> [Titanic — Tidymodels Tour](../index.qmd). The four base-model documents
> ([XGBoost](02-xgboost.qmd), [Random Forest](03-random-forest.qmd),
> [Neural Net](04-neural-net.qmd), [Logistic Regression](06-logistic.qmd)) each
> saved their *tuning results* (every candidate model and its out-of-fold
> predictions) to `models/<model>_res.rds`.
> Here we feed all of those candidates to [`stacks`](https://stacks.tidymodels.org/),
> which learns a regularised linear blend over them — usually beating any single
> model, because the members make different mistakes.

## Why stack?

Each model family has a different inductive bias: gradient boosting chases
interactions, the random forest averages decorrelated trees, the neural net
learns a smooth function of the principal components. A stack fits a
*meta-learner* (penalised logistic regression) on the members' out-of-fold
predictions, keeping only the candidates that pull their weight.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

source(here::here("R", "00_setup.R"))   # tidyverse + tidymodels, seed = 42

library(tidyverse)
library(tidymodels)
library(here)
library(stacks)
# The member engines must be attached so fit_members() can refit them.
library(ranger)
library(xgboost)
library(nnet)
library(glmnet)
```

## Load the tuning results

```{r}
#| label: load-candidates
# Each *_res.rds is a tune_grid result saved with control_grid(save_pred = TRUE,
# save_workflow = TRUE) — exactly what stacks needs to assemble candidates.
xgb_res      <- readRDS(here("models", "xgboost_res.rds"))
rf_res       <- readRDS(here("models", "rf_res.rds"))
nnet_res     <- readRDS(here("models", "nnet_res.rds"))
logistic_res <- readRDS(here("models", "logistic_res.rds"))
```

## Build, blend, and fit the ensemble

```{r}
#| label: build-stack
#| message: false
#| warning: false

set.seed(42)
titanic_stack <- stacks() |>
  add_candidates(xgb_res) |>
  add_candidates(rf_res) |>
  add_candidates(nnet_res) |>
  add_candidates(logistic_res) |>
  blend_predictions() |>     # LASSO over candidates -> keeps the useful ones
  fit_members()              # refit the retained members on all training data

titanic_stack
```

### Which members survived the blend?

```{r}
#| label: weights
#| fig-cap: "Blended stacking coefficients — the members the meta-learner kept."
#| fig-width: 8
#| fig-height: 5
autoplot(titanic_stack, type = "weights") +
  scale_colour_manual(values = grDevices::colorRampPalette(kakapo_pal)(10)) +
  scale_fill_manual(values = grDevices::colorRampPalette(kakapo_pal)(10))
```

The neural net typically earns the largest single weight, with several boosted
trees and a random forest rounding out the blend — different biases, combined.

## Save the fitted ensemble

```{r}
#| label: save
saveRDS(titanic_stack, here("models", "stack_fit.rds"))
```

## Build the Kaggle submission

```{r}
#| label: submit
test <- read_csv(here("data", "processed", "test.csv"), show_col_types = FALSE)

stack_pred <- predict(titanic_stack, new_data = test, type = "class")

submission <- tibble(
  PassengerId = test$PassengerId,
  Survived    = as.integer(as.character(stack_pred$.pred_class))
)
write_csv(submission, here("models", "submission_stack.csv"))

submission |> count(Survived) |> mutate(prop = n / sum(n))
```

Upload it the same way as any other model:

```bash
kaggle competitions submit -c titanic -f models/submission_stack.csv \
  -m "stacked ensemble (xgb + rf + nnet) via stacks"
```

## Next steps

- The parameterized [submission doc](submission.qmd) also accepts
  `-P model:stack`.
- Compare the stack's leaderboard score against the individual models on the
  [master document](../index.qmd) — does the blend actually win on the held-out
  Kaggle test set, or only in cross-validation?
- See the [Kaggle setup guide](setup-kaggle.md) for authentication.
