---
title: "Model 5 — Logistic Regression"
subtitle: "A penalised linear baseline with glmnet — interpretable and fast"
---

> **Where we are:** A fourth base learner for the
> [Titanic — Tidymodels Tour](../index.qmd), following the same
> recipe → spec → workflow → tune → finalize → evaluate → save structure as the
> [XGBoost](02-xgboost.qmd), [Random Forest](03-random-forest.qmd) and
> [Neural Net](04-neural-net.qmd) docs. It also feeds the
> [stacked ensemble](05-stack.qmd).

## Why logistic regression?

It is the classic starting point for binary classification: a linear model of the
log-odds of survival. With the `glmnet` engine we get **elastic-net
regularisation** — `penalty` (how much to shrink) and `mixture` (ridge ↔ lasso) —
which guards against overfitting the wide, PCA-augmented feature space and does
automatic feature selection. It is fast, interpretable (signed coefficients), and
a strong, honest baseline the fancier models must beat.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

source(here::here("R", "00_setup.R"))      # tidyverse + tidymodels, seed = 42
source(here("R", "02_features.R"))         # make_recipe(data)

library(tidyverse)
library(tidymodels)
library(here)
library(glmnet)   # penalised logistic regression engine
```

```{r}
#| label: load-data
#| message: false

train <- read_csv(here("data", "processed", "train.csv")) |>
  mutate(Survived = factor(Survived, levels = c("0", "1")))
```

## The recipe

The same shared recipe as every other model — a fair comparison on identical
features (including the principal components).

```{r}
#| label: recipe
logistic_recipe <- make_recipe(train)
logistic_recipe
```

## The model specification

```{r}
#| label: spec
# Elastic-net logistic regression: penalty = shrinkage strength, mixture = the
# ridge(0) <-> lasso(1) balance. Both tuned.
logistic_spec <- logistic_reg(
  penalty = tune(),
  mixture = tune()
) |>
  set_engine("glmnet") |>
  set_mode("classification")

logistic_spec
```

## The workflow

```{r}
#| label: workflow
logistic_wf <- workflow() |>
  add_recipe(logistic_recipe) |>
  add_model(logistic_spec)
```

## Tuning with cross-validation

```{r}
#| label: resamples
set.seed(42)
folds <- vfold_cv(train, v = 10, strata = Survived)
```

```{r}
#| label: grid
# The recipe adds PCA internally, so num_comp is tunable too. finalize() against
# a prepped predictor set resolves any data-dependent ranges.
preds <- prep(make_recipe(train, num_comp = 2)) |>
  bake(new_data = NULL) |>
  select(-Survived, -PassengerId)

logistic_params <- extract_parameter_set_dials(logistic_wf) |>
  finalize(preds) |>
  update(num_comp = num_comp(range = c(2L, 10L)))

set.seed(42)
logistic_grid <- grid_space_filling(logistic_params, size = 18)
```

```{r}
#| label: tune
#| message: false
#| warning: false

logistic_metrics <- metric_set(roc_auc, accuracy)

set.seed(42)
logistic_tuned <- tune_grid(
  logistic_wf,
  resamples = folds,
  grid      = logistic_grid,
  metrics   = logistic_metrics,
  control   = control_grid(save_pred = TRUE, save_workflow = TRUE, event_level = "second")
)
```

```{r}
#| label: save-res
saveRDS(logistic_tuned, here("models", "logistic_res.rds"))
```

### Tuning results

```{r}
#| label: tune-plot
#| fig-cap: "ROC-AUC across the penalty / mixture grid (autoplot)."
#| fig-width: 8
#| fig-height: 5

autoplot(logistic_tuned, metric = "roc_auc")
```

```{r}
#| label: tune-table
show_best(logistic_tuned, metric = "roc_auc", n = 5)
```

## Finalize and fit on the full training set

```{r}
#| label: finalize
best_logistic <- select_best(logistic_tuned, metric = "roc_auc")
best_logistic

final_logistic_wf <- finalize_workflow(logistic_wf, best_logistic)

set.seed(42)
logistic_fit <- fit(final_logistic_wf, data = train)
```

## Cross-validated performance

```{r}
#| label: cv-metrics
collect_metrics(logistic_tuned) |>
  semi_join(best_logistic, by = names(best_logistic)) |>
  select(.metric, mean, std_err, n)
```

::: {.callout-tip}
These CV metrics feed the results table on the
[master document](../index.qmd) automatically — it reads `models/logistic_res.rds`
at render time, so there's nothing to copy by hand.
:::

### Confusion matrix

```{r}
#| label: confusion
#| fig-cap: "Confusion matrix from out-of-fold CV predictions."
#| fig-width: 5
#| fig-height: 4

logistic_cv_preds <- collect_predictions(logistic_tuned, parameters = best_logistic)

logistic_cv_preds |>
  conf_mat(truth = Survived, estimate = .pred_class) |>
  autoplot(type = "heatmap") +
  scale_fill_gradient(low = "white", high = kakapo_pal[1]) +
  labs(title = "Logistic Regression — CV confusion matrix")
```

### Coefficients (the linear story)

Unlike the tree models, a logistic regression hands us **signed** effects: the
log-odds contribution of each (standardised) feature at the chosen penalty.
Positive pushes toward survival, negative toward not.

```{r}
#| label: coefficients
#| fig-cap: "Largest elastic-net coefficients (log-odds). Green = toward survival."
#| fig-width: 7
#| fig-height: 5

logistic_fit |>
  extract_fit_parsnip() |>
  tidy() |>
  filter(term != "(Intercept)", estimate != 0) |>
  slice_max(abs(estimate), n = 15) |>
  ggplot(aes(x = reorder(term, abs(estimate)), y = estimate,
             fill = estimate > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c(`TRUE` = kakapo_pal[1], `FALSE` = kakapo_pal[5]),
                    labels = c(`TRUE` = "toward survival", `FALSE` = "toward death"),
                    name = NULL) +
  labs(title = "Logistic Regression coefficients",
       x = NULL, y = "Coefficient (log-odds)") +
  theme_minimal()
```

The same engineered signals — `GroupSurvival`, `Title`, `Sex`, class — carry the
weight, now with an explicit direction.

## Save the fitted workflow

```{r}
#| label: save
saveRDS(logistic_fit, file = here("models", "logistic_fit.rds"))
```

## Submission

```bash
quarto render docs/submission.qmd -P model:logistic
```

Loads `models/logistic_fit.rds`, predicts on the processed test set, writes
`models/submission_logistic.csv`, and prints the `kaggle competitions submit`
command. See the [Kaggle setup guide](setup-kaggle.md).

## Next steps

- See how it blends in the **[Stacked Ensemble](05-stack.qmd)**.
- Compare CV scores on the **[master results table](../index.qmd)**.
