---
title: "Model 6 — Support Vector Machine"
subtitle: "A radial-basis-function SVM with kernlab — a nonlinear kernel boundary"
---

> **Where we are:** Another base learner for the
> [Titanic — Tidymodels Tour](../index.qmd), same
> recipe → spec → workflow → tune → finalize → evaluate → save structure as the
> other model docs, and another candidate for the
> [stacked ensemble](05-stack.qmd).

## Why an SVM?

A support vector machine with a **radial basis function (RBF)** kernel draws a
smooth, curved decision boundary in a high-dimensional feature space — a
genuinely different shape of model from the trees, the linear logistic fit, and
the small neural net. That difference is exactly what a stack likes: when the
SVM is wrong, it tends to be wrong on *different* passengers. We tune `cost`
(margin softness) and `rbf_sigma` (kernel width).

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

source(here::here("R", "00_setup.R"))
source(here("R", "02_features.R"))

library(tidyverse)
library(tidymodels)
library(here)
library(kernlab)   # the SVM engine
```

```{r}
#| label: load-data
#| message: false

train <- read_csv(here("data", "processed", "train.csv")) |>
  mutate(Survived = factor(Survived, levels = c("0", "1")))
```

## The recipe

The shared recipe — normalisation matters a lot for a distance-based kernel, and
`make_recipe()` already normalises before adding the principal components.

```{r}
#| label: recipe
svm_recipe <- make_recipe(train)
svm_recipe
```

## The model specification

```{r}
#| label: spec
svm_spec <- svm_rbf(
  cost      = tune(),
  rbf_sigma = tune()
) |>
  set_engine("kernlab") |>
  set_mode("classification")

svm_spec
```

## The workflow

```{r}
#| label: workflow
svm_wf <- workflow() |>
  add_recipe(svm_recipe) |>
  add_model(svm_spec)
```

## Tuning with cross-validation

```{r}
#| label: resamples
set.seed(42)
folds <- vfold_cv(train, v = 10, strata = Survived)
```

```{r}
#| label: grid
preds <- prep(make_recipe(train, num_comp = 2)) |>
  bake(new_data = NULL) |>
  select(-Survived, -PassengerId)

svm_params <- extract_parameter_set_dials(svm_wf) |>
  finalize(preds) |>
  update(num_comp = num_comp(range = c(2L, 10L)))

set.seed(42)
svm_grid <- grid_space_filling(svm_params, size = 18)
```

```{r}
#| label: tune
#| message: false
#| warning: false

svm_metrics <- metric_set(roc_auc, accuracy)

set.seed(42)
svm_tuned <- tune_grid(
  svm_wf,
  resamples = folds,
  grid      = svm_grid,
  metrics   = svm_metrics,
  control   = control_grid(save_pred = TRUE, save_workflow = TRUE, event_level = "second")
)
```

```{r}
#| label: save-res
saveRDS(svm_tuned, here("models", "svm_res.rds"))
```

### Tuning results

```{r}
#| label: tune-plot
#| fig-cap: "ROC-AUC across the cost / rbf_sigma grid (autoplot)."
#| fig-width: 8
#| fig-height: 5

autoplot(svm_tuned, metric = "roc_auc")
```

```{r}
#| label: tune-table
show_best(svm_tuned, metric = "roc_auc", n = 5)
```

## Finalize and fit on the full training set

```{r}
#| label: finalize
best_svm <- select_best(svm_tuned, metric = "roc_auc")
best_svm

final_svm_wf <- finalize_workflow(svm_wf, best_svm)

set.seed(42)
svm_fit <- fit(final_svm_wf, data = train)
```

## Cross-validated performance

```{r}
#| label: cv-metrics
collect_metrics(svm_tuned) |>
  semi_join(best_svm, by = names(best_svm)) |>
  select(.metric, mean, std_err, n)
```

::: {.callout-tip}
These CV metrics feed the results table on the
[master document](../index.qmd) automatically — it reads `models/svm_res.rds`
at render time, so there's nothing to copy by hand.
:::

### Confusion matrix

```{r}
#| label: confusion
#| fig-cap: "Confusion matrix from out-of-fold CV predictions."
#| fig-width: 5
#| fig-height: 4

svm_cv_preds <- collect_predictions(svm_tuned, parameters = best_svm)

svm_cv_preds |>
  conf_mat(truth = Survived, estimate = .pred_class) |>
  autoplot(type = "heatmap") +
  scale_fill_gradient(low = "white", high = kakapo_pal[1]) +
  labs(title = "SVM (RBF) — CV confusion matrix")
```

::: {.callout-note}
## No native variable importance
An RBF-kernel SVM is effectively a black box — there is no coefficient or split
to read importance from. If you want it, use a *model-agnostic* permutation
approach (e.g. `DALEX`/`vip::vi(method = "permute")`). We lean on the other
models for the importance story.
:::

## Save the fitted workflow

```{r}
#| label: save
saveRDS(svm_fit, file = here("models", "svm_fit.rds"))
```

## Submission

```bash
quarto render docs/submission.qmd -P model:svm
```

Writes `models/submission_svm.csv`. See the [Kaggle setup guide](setup-kaggle.md).

## Next steps

- See how it blends in the **[Stacked Ensemble](05-stack.qmd)**.
- Compare CV scores on the **[master results table](../index.qmd)**.
