---
title: "Model 7 — MARS"
subtitle: "Multivariate adaptive regression splines with earth — nonlinear, interpretable"
---

> **Where we are:** The last individual base learner in the
> [Titanic — Tidymodels Tour](../index.qmd), same structure as the other model
> docs and a final candidate for the [stacked ensemble](05-stack.qmd).

## Why MARS?

**Multivariate Adaptive Regression Splines** fit the response as a sum of hinge
functions — piecewise-linear pieces that bend at data-driven knots. That lets it
capture *nonlinear* effects (a survival rate that changes slope at a certain age
or fare) and simple interactions, while staying more interpretable than a kernel
machine. Its bias sits between the linear logistic model and the trees — useful
diversity for the stack. We tune `num_terms` (model size) and `prod_degree`
(1 = additive, 2 = allow pairwise interactions).

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

source(here::here("R", "00_setup.R"))
source(here("R", "02_features.R"))

library(tidyverse)
library(tidymodels)
library(here)
library(earth)   # the MARS engine
library(vip)     # earth exposes variable importance
```

```{r}
#| label: load-data
#| message: false

train <- read_csv(here("data", "processed", "train.csv")) |>
  mutate(Survived = factor(Survived, levels = c("0", "1")))
```

## The recipe

```{r}
#| label: recipe
mars_recipe <- make_recipe(train)
mars_recipe
```

## The model specification

```{r}
#| label: spec
mars_spec <- mars(
  num_terms   = tune(),
  prod_degree = tune()
) |>
  set_engine("earth") |>
  set_mode("classification")

mars_spec
```

## The workflow

```{r}
#| label: workflow
mars_wf <- workflow() |>
  add_recipe(mars_recipe) |>
  add_model(mars_spec)
```

## Tuning with cross-validation

```{r}
#| label: resamples
set.seed(42)
folds <- vfold_cv(train, v = 10, strata = Survived)
```

```{r}
#| label: grid
preds <- prep(make_recipe(train, num_comp = 2)) |>
  bake(new_data = NULL) |>
  select(-Survived, -PassengerId)

mars_params <- extract_parameter_set_dials(mars_wf) |>
  finalize(preds) |>
  update(num_comp = num_comp(range = c(2L, 10L)))

set.seed(42)
mars_grid <- grid_space_filling(mars_params, size = 18)
```

```{r}
#| label: tune
#| message: false
#| warning: false

mars_metrics <- metric_set(roc_auc, accuracy)

set.seed(42)
mars_tuned <- tune_grid(
  mars_wf,
  resamples = folds,
  grid      = mars_grid,
  metrics   = mars_metrics,
  control   = control_grid(save_pred = TRUE, save_workflow = TRUE, event_level = "second")
)
```

```{r}
#| label: save-res
saveRDS(mars_tuned, here("models", "mars_res.rds"))
```

### Tuning results

```{r}
#| label: tune-plot
#| fig-cap: "ROC-AUC across the num_terms / prod_degree grid (autoplot)."
#| fig-width: 8
#| fig-height: 5

autoplot(mars_tuned, metric = "roc_auc")
```

```{r}
#| label: tune-table
show_best(mars_tuned, metric = "roc_auc", n = 5)
```

## Finalize and fit on the full training set

```{r}
#| label: finalize
best_mars <- select_best(mars_tuned, metric = "roc_auc")
best_mars

final_mars_wf <- finalize_workflow(mars_wf, best_mars)

set.seed(42)
mars_fit <- fit(final_mars_wf, data = train)
```

## Cross-validated performance

```{r}
#| label: cv-metrics
collect_metrics(mars_tuned) |>
  semi_join(best_mars, by = names(best_mars)) |>
  select(.metric, mean, std_err, n)
```

::: {.callout-tip}
These CV metrics feed the results table on the
[master document](../index.qmd) automatically — it reads `models/mars_res.rds`
at render time, so there's nothing to copy by hand.
:::

### Confusion matrix

```{r}
#| label: confusion
#| fig-cap: "Confusion matrix from out-of-fold CV predictions."
#| fig-width: 5
#| fig-height: 4

mars_cv_preds <- collect_predictions(mars_tuned, parameters = best_mars)

mars_cv_preds |>
  conf_mat(truth = Survived, estimate = .pred_class) |>
  autoplot(type = "heatmap") +
  scale_fill_gradient(low = "white", high = kakapo_pal[1]) +
  labs(title = "MARS — CV confusion matrix")
```

### Variable importance

MARS reports how often each feature enters the retained hinge terms (the `earth`
`nsubsets` criterion) — the features the splines actually bend on.

```{r}
#| label: vip
#| fig-cap: "MARS variable importance (number of subsets each feature appears in)."
#| fig-width: 7
#| fig-height: 5

mars_fit |>
  extract_fit_parsnip() |>
  vip(num_features = 15, aesthetics = list(fill = kakapo_pal[1])) +
  labs(title = "MARS variable importance") +
  theme_minimal()
```

## Save the fitted workflow

```{r}
#| label: save
saveRDS(mars_fit, file = here("models", "mars_fit.rds"))
```

## Submission

```bash
quarto render docs/submission.qmd -P model:mars
```

Writes `models/submission_mars.csv`. See the [Kaggle setup guide](setup-kaggle.md).

## Next steps

- See how it blends in the **[Stacked Ensemble](05-stack.qmd)**.
- Compare CV scores on the **[master results table](../index.qmd)**.
