---
title: "Building a Kaggle Submission"
subtitle: "One parameterized document, three models, leaderboard-ready CSVs"
params:
  model: "xgboost"
---

> **Where we are:** The final stop on the
> [Titanic — Tidymodels Tour](../index.qmd). Each model document
> ([XGBoost](02-xgboost.qmd), [Random Forest](03-random-forest.qmd),
> [Neural Net](04-neural-net.qmd)) fits a workflow and saves it to
> `models/<model>_fit.rds`. This document is **parameterized**: tell it which
> model, and it loads that fit, predicts on the test set, and writes a
> Kaggle-ready CSV.

## How to render this document per model

This `.qmd` has a `model` parameter (default `"xgboost"`). Override it on the
command line with `-P model:<value>`:

```bash
# XGBoost (also the default if you omit -P)
quarto render docs/submission.qmd -P model:xgboost

# Random forest
quarto render docs/submission.qmd -P model:rf

# Neural net
quarto render docs/submission.qmd -P model:nnet
```

Each render reads the matching `models/<model>_fit.rds`, writes
`models/submission_<model>.csv`, and prints the exact upload command. You must
have fitted (rendered) the corresponding model document first so the `.rds`
exists.

## Setup

```{r}
#| label: setup
#| message: false
#| warning: false

# We need the tidymodels stack to load the saved workflow and call predict().
# 00_setup.R also sets paths and the seed.
source(here::here("R", "00_setup.R"))
library(readr)
```

## Resolve and validate the requested model

```{r}
#| label: resolve-model
# The parameter comes in from the YAML default or the -P flag on the CLI.
requested_model <- params$model

# Map each valid model key to its saved fit. This single lookup table is the
# source of truth for which models are submittable.
model_paths <- c(
  xgboost = here::here("models", "xgboost_fit.rds"),
  rf      = here::here("models", "rf_fit.rds"),
  nnet    = here::here("models", "nnet_fit.rds")
)

# Fail loudly and helpfully if someone passes an unknown model.
if (!requested_model %in% names(model_paths)) {
  stop(
    "Unknown model '", requested_model, "'. ",
    "Choose one of: ", paste(names(model_paths), collapse = ", "), ". ",
    "Example: quarto render docs/submission.qmd -P model:rf",
    call. = FALSE
  )
}

fit_path <- model_paths[[requested_model]]

# Make sure the fitted workflow actually exists — i.e. the model doc was
# rendered first.
if (!file.exists(fit_path)) {
  stop(
    "Fitted model not found at '", fit_path, "'. ",
    "Render the matching model document first (e.g. quarto render ",
    "docs/02-xgboost.qmd) so the .rds is saved.",
    call. = FALSE
  )
}

message("Building submission for model: ", requested_model)
message("Loading fitted workflow from: ", fit_path)
```

We are building a submission for **`r requested_model`**.

## Load the model and the test data

```{r}
#| label: load
# The saved object is a fully fitted workflow (recipe + model), so it knows how
# to preprocess raw test data exactly as it was trained.
model_fit <- readRDS(fit_path)

# The test set has the same columns as train EXCEPT Survived — that's what we
# predict. PassengerId is what Kaggle keys submissions on.
test <- readr::read_csv(here::here("data", "raw", "test.csv"))

dplyr::glimpse(test)
```

## Predict and assemble the submission

```{r}
#| label: predict
# predict() runs the workflow's recipe on the raw test data, then the model.
# type = "class" returns the hard 0/1 label Kaggle wants.
preds <- predict(model_fit, new_data = test, type = "class")

# Kaggle requires EXACTLY two columns: PassengerId and Survived, with Survived
# as 0/1 integers. .pred_class is a factor ("0"/"1"); convert via character so
# we don't accidentally get factor codes (1/2).
submission <- tibble::tibble(
  PassengerId = test$PassengerId,
  Survived    = as.integer(as.character(preds$.pred_class))
)

# Sanity checks: right number of rows, right value set, no missing predictions.
stopifnot(
  nrow(submission) == nrow(test),
  all(submission$Survived %in% c(0L, 1L)),
  !anyNA(submission$Survived)
)

head(submission, 10)
```

## Write the CSV

```{r}
#| label: write
out_path <- here::here("models", paste0("submission_", requested_model, ".csv"))

readr::write_csv(submission, out_path)

message("Wrote ", nrow(submission), " predictions to: ", out_path)
```

A quick summary of the predicted survival split (a sanity check against the
~38% base rate seen in the [EDA](01-eda.qmd)):

```{r}
#| label: summary
submission |>
  dplyr::count(Survived) |>
  dplyr::mutate(prop = n / sum(n))
```

## Submit to Kaggle

Push the CSV to the leaderboard with the Kaggle CLI. The command below is built
for the model you rendered:

```{r}
#| label: kaggle-cmd
#| echo: false
#| results: asis

cmd <- sprintf(
  'kaggle competitions submit -c titanic -f models/submission_%s.csv -m "%s via tidymodels"',
  requested_model, requested_model
)
cat("```bash\n", cmd, "\n```\n", sep = "")
```

::: {.callout-note}
## First time? Authenticate first
The `kaggle` CLI needs your API token (`~/.kaggle/kaggle.json`) before any
`submit` or `download` will work. The
[Kaggle setup guide](setup-kaggle.md) walks through creating and installing it.
:::

::: {.callout-tip}
## Daily submission limit
Kaggle caps Titanic submissions per day (commonly ~10). Submit your *best*
candidate by CV score rather than spraying every experiment at the leaderboard.
:::

## Next steps

- Compare each model's leaderboard score against its **CV score** recorded in
  the [master results table](../index.qmd) — divergence is a lesson in
  generalization.
- Re-render for the other models:
  `-P model:xgboost`, `-P model:rf`, `-P model:nnet`.
- Need auth help? See the **[Kaggle setup guide](setup-kaggle.md)**.
- Back to the **[master document](../index.qmd)** for the whole-project picture.
